The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-06 rational.tex
related work

@@

\paragraph*{Declarative Continuations.} 
In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which values and continuations are treated as opposites. This is essentially
what we are proposing with one fundamental difference: our underlying
language is not the $\lambda$-calculus but a language of pure isomorphisms in
which information is preserved. This shift of perspective enables us to
distill and generalize the duality of values and continuations: in
particular, in the conventional $\lambda$-calculus setting values and
continuations can be erased and duplicated which makes it difficult to
maintain the correspondence between a value and its negative counterpart. In
contrast, in our setting, one can start from the empty type $0$, introduce a
positive value and its negative counterpart, and let each of these flow in
arbitrary ways. The entire framework guarantees that neither the value nor
its negative counterpart will be deleted or duplicated and hence that, in any
closed program, the ``debt'' corresponding to the negative value is paid off
exactly once. The forward and backward executions in our framework correspond
to call-by-value and call-by-name. This duality was observed by Filinski and
others following him but it is particularly clean in our framework.

\paragraph*{The Duality of Computation.}
Wadler the reloaded paper, does not consider the subtraction type because its
``computational interpretation is not familiar.'' Curien and Herbelin study
duality in classical logic, show that it exchanges call-by-value and
call-by-name. They extend classical natural deduction with subtraction but
give it no computational meaning. Crolard (in the formulae-as-types
interpretation of subtractive logic) address the computational interpretation
of subtraction. He explains the type $A-B$ as the of \emph{coroutines} with a
local environment of type $A$ and a continuation of type $B$. The description
is complicated by what is essentially the desire to enforce linearity
constraints so that coroutines cannot access the local environment of other
coroutines. Must cite Selinger control categories in this context of duality
but I am not sure what to say: it assumes cartesian closed categories for one
thing and not symmetric monoidal ones. Ariola, Herbelin, and Sabry also use
subtractive types to explain delimited continuations.

\paragraph*{Polarized Logic.} Polarized logic is the logic of values and
continuations: each connective can be given a positive or negative polarity,
which can be related to evaluation
order~\cite{10.1109/LICS.2010.23}. Negative types in this work are a mix of
our negative and fractional types: they include a notion of demand which
corresponds to our negative types and a notion of pattern-matching on
continuations (i.e., a notion of constraints on the context) which
corresponds to our fractional types.

@@

\newpage

Curien and Herbelin~\cite{Curien:2000} introduce a calculus that exhibits
symmetries between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. which whose computational interpretation is left for future work. Curien
and Herbelin informally explain values of types $A-B$ either as a context
expecting a function $A \rightarrow B$ or as a \emph{pair} containing a value
and a continuation. In our interpretation, a value of type $A-B$ is
\emph{either} a value of $A$ or a demand for a value of type $B$.

@@

Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
logic which contains a dual to implication. Her work has been distilled in
the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.

In a seemingly unrelated development, Girard in 1987 introduced linear

@@

Connections to type logical grammars and Lambek calculus

Some background on {{langRev}} because it is mentioned in the next section.

>>>>

\paragraph*{Declarative Continuations.} 
In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which values and continuations are treated as opposites. In more detail, the
symmetric calculus contains a ``value'' fragment and a ``continuation''
fragment which are mirror images. Pairs and sums are treated as duals in the
sense that the ``value'' fragment includes pairs whose mirror image in the
``continuation'' fragment are sums. In contrast, our language includes pairs
and sums in the value fragment and two symmetries: one that maps the pairs to
fractions and another that maps the sums to subtractions.

\paragraph*{The Duality of Computation.}
The precise duality between call-by-name and call-by-value was further
investigated by Selinger~\cite{Selinger:2001:CCD:966910.966911}. Curien and
Herbelin~\cite{Curien:2000} introduce a calculus that exhibits symmetries
between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. This is to be contrasted with our interpretation of a value of that type
as \emph{either} a value of $A$ or a demand for a value of type $B$. This
calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
interpretation to the subtraction connective and like the original symmetric
calculus of Filinksi, introduces a duality that relates sums to products and
vice-versa.

\paragraph*{Subtractive Logic.} 
Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
logic which contains a dual to implication. Her work has been distilled in
the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
Crolard explains the type $A-B$ as the of \emph{coroutines} with a local
environment of type $A$ and a continuation of type $B$. The description is
complicated by what is essentially the desire to enforce linearity
constraints so that coroutines cannot access the local environment of other
coroutines.

@@

@@

@@

@@

This is essentially what we are proposing with one fundamental difference:
our underlying language is not the $\lambda$-calculus but a language of pure
isomorphisms in which information is preserved. This shift of perspective
enables us to distill and generalize the duality of values and continuations:
in particular, in the conventional $\lambda$-calculus setting values and
continuations can be erased and duplicated which makes it difficult to
maintain the correspondence between a value and its negative counterpart. In
contrast, in our setting, one can start from the empty type $0$, introduce a
positive value and its negative counterpart, and let each of these flow in
arbitrary ways. The entire framework guarantees that neither the value nor
its negative counterpart will be deleted or duplicated and hence that, in any
closed program, the ``debt'' corresponding to the negative value is paid off
exactly once. The forward and backward executions in our framework correspond
to call-by-value and call-by-name. This duality was observed by Filinski and
others following him but it is particularly clean in our framework.
